Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
more

\theoremstyle{definition}
\newtheorem{definition}{Definition}%[section]
\newtheorem{example}{Example}%[section]

\theoremstyle{remark}
\newtheorem*{proofsketch}{Proof sketch}

@@

The translation $\conv$ defined in Figure~\ref{fig:A} translates the $\Sigma$-formula
$\varphi$ into an $\extendsig{\Sigma}$-formula. It relies
on the auxiliary function $\conv_0$, which takes two arguments:\ the term $t$
to translate and a \relax{polarity}~$p$, which is either $\ppos$, $\pneg$, or
$\pnone$ (none). \vthinspace$\conv_0$ returns a pair $( t'\negvthinspace,\, \chi )$, where $t'$ is a term of
type $\tau$ and $\chi$ is an additional constraint.
This constraint is needed if the polarity is $\pnone$; otherwise, it is encoded
directly in $t'$ and $\chi$ is simply $\ltrue$.
In the auxiliary function $\conv_0$, the function $\pol(
\boolop, i, p )$ returns the polarity of the $i$\vvthinspace{th} argument of an
application of~$\boolop$ having polarity $p$, where $\boolop$ is an interpreted
predicate.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. 
This formula is unsatisfiable only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{\con{s}}$ can be
interpreted as the integers and $\farg{\con{s}}$ as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable
only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x.\;\, \varphi_\con{s} \}$ is admissible,
where $\varphi_\con{s}$ is the formula $\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr)$.
%\ct{
%You cannot possibly mean "if and only if" in the two sentences above.
%Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
%(1) is sat iff (2) is sat. That is immediate.
%}
% You're right, changed both to "only if" --ajr
%As stated
%in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
%formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
%function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
%model.
%The rightmost conjunct $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$ 
%ensures that if $\con{s}( \con{c} ) > 100$ is a constraint that 
%formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
%satisfies $\varphi_{\con{s}}[\con{c}/x]$. 
%The nested existential
%$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
%a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
%any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. 
It is easy to see that any interpretation $\I$ satisfying formula~(\ref{eq:ex-after}) satisfies $L_0 \cup \absconstraints( L_0 )$,
where $L_0 = \{ \con{s}( \con{c} ) > 100 \}$, 
where $\absconstraints( L_0 )$ is the set consisting of the single constraint $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$.
%The latter ensures that $\I$ satisfies a formula that entails both $\con{s}( \con{c} ) > 100$ and $\varphi_{\con{s}}[\con{c}/x]$.
Since $\I$ also satisfies both the translated function definition for $\con{s}$---i.e.,
the first conjunct of formula~(\ref{eq:ex-after})---and $\absconstraints( L_0
)$, it must also satisfy the formula $\psi_1$:
\[
\lite\bigl( \con{c} \leq 0,\; 
            \con{s}(\con{c}) \teq 0,\;
            \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 )
            \wedge \exists b : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( b ) \teq \con{c}-1 \bigr)
\]
The existential constraint in the above formula
ensures that whenever $\I$ satisfies the set $L_1 = L_0 \cup \{ \neg \con{c} \leq 0,\; \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 ) \}$,
$\I$ satisfies $\absconstraints( L_1 )$ as well.
Hence, by repeated application of this reasoning, it follows that
a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy the formula $\psi$:
%
\[\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!n-1} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq \con{c}-i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-n \leq 0 \wedge \con{s}( \con{c}-n ) \teq 0
\end{split}\]
%
This formula is closed under function expansion
since it entails $\varphi_\con{s}[(\con{c}-i)/x]$ for $i = 0, \ldots, n$,
and it contains only $\con{s}$ applications corresponding to $\con{s}( \con{c}-i )$ for $i = 0, \ldots, n$.
Since $\{\forallf{\con{s}} x.\;\, \varphi_\con{s}\}$ is admissible,
there exists a $\Sigma$-interpretation satisfying $\psi \wedge \forallf{\con{s}} x.\;\, \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

In this section, we evaluate both the overall impact of the translation
introduced in the previous section and the performance of individual SMT
techniques. The experimental data is publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/SMT2015-recfun/}}

>>>>

\theoremstyle{definition}
\newtheorem{definition}{Definition}%[section]

\theoremstyle{remark}
\newtheorem*{proofsketch}{Proof sketch}
\newtheorem{example}{Example}%[section]

@@

The translation $\conv$ defined in Figure~\ref{fig:A} translates the $\Sigma$-formula
$\varphi$ into an $\extendsig{\Sigma}$-formula. It relies
on the auxiliary function $\conv_0$, which takes two arguments:\ the term $t$
to translate and a \relax{polarity}~$p$, which is either $\ppos$, $\pneg$, or
$\pnone$ (none). \vthinspace$\conv_0$ returns a pair $( t'\negvthinspace,\, \chi )$, where $t'$ is a term of
type $\tau$ and $\chi$ is an additional constraint.
This constraint is needed if the polarity is $\pnone$; otherwise, it is encoded
directly in $t'$ and $\chi$ is simply $\ltrue$.
In the auxiliary function $\conv_0$, the function $\pol(
\boolop, i, p )$ returns the polarity of the $i$\vvthinspace{th} argument of an
application of~$\boolop$ having polarity $p$ if $\boolop$ is an interpreted
predicate or $\pnone$ otherwise.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. 
This formula is unsatisfiable only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{\con{s}}$ can be
interpreted as the integers and $\farg{\con{s}}$ as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable
only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x.\;\, \varphi_\con{s} \}$ is admissible,
where $\varphi_\con{s}$ is the formula $\lite\bigl( x \leq 0,\; \con{s}(x) \teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr)$.
%\ct{
%You cannot possibly mean "if and only if" in the two sentences above.
%Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
%(1) is sat iff (2) is sat. That is immediate.
%}
% You're right, changed both to "only if" --ajr
%As stated
%in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
%formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
%function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
%model.
%The rightmost conjunct $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$ 
%ensures that if $\con{s}( \con{c} ) > 100$ is a constraint that 
%formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
%satisfies $\varphi_{\con{s}}[\con{c}/x]$. 
%The nested existential
%$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
%a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
%any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. 
It is easy to see that any interpretation $\I$ satisfying formula~(\ref{eq:ex-after}) satisfies $L_0 \cup \absconstraints( L_0 )$,
where $L_0 = \{ \con{s}( \con{c} ) > 100 \}$, 
where $\absconstraints( L_0 )$ is the set consisting of the single constraint $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$.
%The latter ensures that $\I$ satisfies a formula that entails both $\con{s}( \con{c} ) > 100$ and $\varphi_{\con{s}}[\con{c}/x]$.
Since $\I$ also satisfies both the translated function definition for $\con{s}$---i.e.,
the first conjunct of formula~(\ref{eq:ex-after})---and $\absconstraints( L_0
)$, it must also satisfy the formula $\psi_1$:
\[
\lite\bigl( \con{c} \leq 0,\; 
            \con{s}(\con{c}) \teq 0,\;
            \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 )
            \wedge \exists b : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( b ) \teq \con{c}-1 \bigr)
\]
The existential constraint in the above formula
ensures that whenever $\I$ satisfies the set $L_1 = L_0 \cup \{ \neg \con{c} \leq 0,\; \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 ) \}$,
$\I$ satisfies $\absconstraints( L_1 )$ as well.
Hence, by repeated application of this reasoning, it follows that
a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy the formula $\psi$:
%
\[\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!n-1} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq \con{c}-i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-n \leq 0 \wedge \con{s}( \con{c}-n ) \teq 0
\end{split}\]
%
This formula is closed under function expansion
since it entails $\varphi_\con{s}[(\con{c}-i)/x]$ for $i = 0, \ldots, n$,
and it contains only $\con{s}$ applications corresponding to $\con{s}( \con{c}-i )$ for $i = 0, \ldots, n$.
Since $\{\forallf{\con{s}} x.\;\, \varphi_\con{s}\}$ is admissible,
there exists a $\Sigma$-interpretation satisfying $\psi \wedge \forallf{\con{s}} x.\;\, \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

In this section, we evaluate both the overall impact of the translation
introduced in the previous section and the performance of individual SMT
techniques. The experimental data is publicly available online.%
\footnote{\url{http://lara.epfl.ch/~reynolds/SMT2015-recfun/}}
